---
title: "Data 621 HW1"
author: "Mikhail Groysman"
date: "February 21, 2019"
output:
  bookdown::html_document2:
    theme: cerulean
    toc: true
    toc_depth: 5
    #code_folding: hide
---

#Data Exploration

```{r, fig.height = 5, fig.width = 10}

ourdata<-read.csv(file="C:/HW/moneyball-training-data.csv")

# visual observation of top records

head(ourdata)

library(knitr)

# mean/median/IQR/NA

kable(summary(ourdata))

```

```{r}

# histogram/density

# near-normal

par(mar=c(2,2,2,2))

par(mfrow = c(2,2))

hist(ourdata$TARGET_WINS)

plot(density(ourdata$TARGET_WINS))

# near-normal, long right-tail, potential outliers

hist(ourdata$TEAM_BATTING_H)

plot(density(ourdata$TEAM_BATTING_H))

# near-normal, some weirdness around the mean

hist(ourdata$TEAM_BATTING_2B)

plot(density(ourdata$TEAM_BATTING_2B))

# heavy right skewed, outliers(?)

hist(ourdata$TEAM_BATTING_3B)

plot(density(ourdata$TEAM_BATTING_3B))

# looks very weird, does not look right, potentially needs to be dropped

hist(ourdata$TEAM_BATTING_HR)

plot(density(ourdata$TEAM_BATTING_HR))

# left skewed, outliers(?)

hist(ourdata$TEAM_BATTING_BB)

plot(density(ourdata$TEAM_BATTING_BB))

# weird, potentially to be dropped

hist(ourdata$TEAM_BATTING_SO)

plot(density(ourdata$TEAM_BATTING_SO,na.rm = TRUE))

# right skewed, outliers(?)

hist(ourdata$TEAM_BASERUN_SB)

plot(density(ourdata$TEAM_BASERUN_SB,na.rm = TRUE))

# right skewed, outliers(?)

hist(ourdata$TEAM_BASERUN_CS)

plot(density(ourdata$TEAM_BASERUN_CS,na.rm = TRUE))

# right skewed, outliers(?)

hist(ourdata$TEAM_FIELDING_E)

plot(density(ourdata$TEAM_FIELDING_E,na.rm = TRUE))

# near-normal, slightly left skewed

hist(ourdata$TEAM_FIELDING_DP)

plot(density(ourdata$TEAM_FIELDING_DP,na.rm = TRUE))

# right skewed, outliers

hist(ourdata$TEAM_PITCHING_BB)

plot(density(ourdata$TEAM_PITCHING_BB,na.rm = TRUE))

# severily right skewed, to be dropped(?)

hist(ourdata$TEAM_PITCHING_H)

plot(density(ourdata$TEAM_PITCHING_H,na.rm = TRUE))

# weird, to be dropped(?)

hist(ourdata$TEAM_PITCHING_HR)

plot(density(ourdata$TEAM_PITCHING_HR,na.rm = TRUE))

# severily right skewed, to be dropped(?)

hist(ourdata$TEAM_PITCHING_SO)

plot(density(ourdata$TEAM_PITCHING_SO,na.rm = TRUE))

```






```{r, fig.height = 12, fig.width = 20}

pairs(ourdata)

```


```{r}

cor(ourdata, use="pairwise.complete.obs", method="pearson") 

```

```{r}

# very high correlation for Home Runs - very suspicious - looks wrong

temp <- lm(TEAM_BATTING_HR~ TEAM_PITCHING_HR, data=ourdata)

with(ourdata,plot(TEAM_PITCHING_HR,TEAM_BATTING_HR))

abline(temp)

```




```{r}

# some data is missing

sum(ourdata$TEAM_BATTING_H)

sum(ourdata$TEAM_PITCHING_H)

sum(ourdata$TEAM_BATTING_BB)

sum(ourdata$TEAM_PITCHING_BB)

ourdata$losses<-162-ourdata$TARGET_WINS

sum(ourdata$TARGET_WINS)

sum(ourdata$losses)

```



Problems with the data

1) Number of wins and number of losses is not the same, so we do not have complete data; which still could be ok for our analysis

2) No strong correlations between dependent and independent variables

3) TEAM_BATTING_HR and TEAM_PITCHING_HR are almost identical , correlation between them 93%. It does not make sense. Their distribution look very weird too. Possible we need to drop both variables.

4) TEAM_PITCHING_H correlates positively, while it should have negative correlation. If we use the variable we need to explain why we use variable that is counter-intuitive.has a lot of outliers. Maximum value is 30,132 or 191 per game, which is not possible (heavily right skewed) It needs to be fixed or dropped.

5) TEAM_BASERUN_CS correlates positively (actually close to zero), while it should have negative correlation. Right-skewed with potential outliers. If we use the variable we need to explain why we use variable that is counter-intuitive.

6) TEAM_BASERUN_HBP has too much missing data. It needs to be dropped.

7) TEAM_PITCHING_BB correlates positively, while it should have negative correlation. If we use the variable we need to explain why we use variable that is counter-intuitive.. It has a lot of outliers, which seem to skew the data. Maybe it could be used after we fix it. 

8) TEAM_PITCHING_SO correlates negatively, while it should have positive correlation. If we use the variable we need to explain why we use variable that is counter-intuitive.. It has a lot of outliers, which seem to skew the data. Maybe it could be used after we fix it.

9) TEAM_FIELDING_DP correlates negatively (close to zero), while it should have positive correlation. If we use the variable we need to explain why we use variable that is counter-intuitive. 

10) Strikeout numbers both for Batting and Pitching look very low. Distributions look weird, skewed.


#Data Preperation

```{r, fixing NA data}

# dropping bad variable

ourdata$TEAM_BATTING_HBP<-NA

# We will keep one dataset the way it is, and we will create a copy of it to manually adjust data

ourdataM<-ourdata

# dropping pitching Home Runs

ourdataM$TEAM_PITCHING_HR<-NA

# dropping outliers

82-(92-71)*2

82+(92-71)*2

# 116 Wins was the maximum, so to use 124 as cut off is very conservative; https://www.mlb.com/news/most-mlb-wins-in-a-season/c-289159676

# The minimum in more resent years was 40; http://writing.jmpressley.net/sports/worstseasons.html

ourdataM<-ourdataM[!(ourdata$TARGET_WINS<40),]

ourdataM<-ourdataM[!(ourdata$TARGET_WINS>124),]

# I use 2018 data as reference; https://www.baseball-reference.com/leagues/MLB/2018.shtml

1454-(1537-1383)*2

1454+(1537-1383)*2

ourdataM<-ourdataM[!(ourdataM$TEAM_BATTING_H<1146),]

ourdataM<-ourdataM[!(ourdataM$TEAM_BATTING_H>1762),]

238-(273-208)*2

238+(273-208)*2

ourdataM<-ourdataM[!(ourdataM$TEAM_BATTING_2B<108),]

ourdataM<-ourdataM[!(ourdataM$TEAM_BATTING_2B>368),]

47-(72-34)*2

47+(72-34)*2

ourdataM<-ourdataM[!(ourdataM$TEAM_BATTING_3B>123),]

# Home Run historical data indicates 3 was the minumum ever; http://www.baseball-almanac.com/recbooks/rb_hr7.shtml

ourdataM<-ourdataM[!(ourdataM$TEAM_BATTING_HR<3),]

512-(580-451)*2

512+(580-451)*2

ourdataM<-ourdataM[!(ourdataM$TEAM_BATTING_BB<254),]

ourdataM<-ourdataM[!(ourdataM$TEAM_BATTING_BB>770),]

# TEAM_BATTING_SO is very inconsistent with the 2018 reference data; I would drop the variable

ourdataM$TEAM_BATTING_SO<-NA

101-(156-56)*2

101+(156-56)*2

ourdataM<-ourdataM[!(ourdata$TEAM_BASERUN_SB>301),]

1518-(1682-1419)*2

1518+(1682-1419)*2

ourdataM<-ourdataM[!(ourdata$TEAM_PITCHING_H<992),]

ourdataM<-ourdataM[!(ourdata$TEAM_PITCHING_H>2044),]

536.5-(611-476)*2

536.5+(611-476)*2

ourdataM<-ourdataM[!(ourdataM$TEAM_PITCHING_BB<266),]

ourdataM<-ourdataM[!(ourdataM$TEAM_PITCHING_BB>807),]

813.5-(968-615)*2

813.5+(968-615)*2

ourdataM<-ourdataM[!(ourdataM$TEAM_PITCHING_SO<107),]

ourdataM<-ourdataM[!(ourdataM$TEAM_PITCHING_SO>1520),]

159-(249.2-127)*2

159+(249.2-127)*2

ourdataM<-ourdataM[!(ourdataM$TEAM_FIELDING_E>404),]

149-(164-131)*2

149+(164-131)*2

ourdataM<-ourdataM[!(ourdataM$TEAM_FIELDING_DP>215),]

ourdataM<-ourdataM[!(ourdataM$TEAM_FIELDING_DP<83),]

ourdataM<-ourdataM[!is.na(ourdataM$INDEX),]

summary(ourdataM)

```

```{r}

# using median to impute missing values

ourdataM$TEAM_PITCHING_SO[is.na(ourdataM$TEAM_PITCHING_SO)]<-827

ourdataM$TEAM_BASERUN_SB[is.na(ourdataM$TEAM_BASERUN_SB)]<-92

ourdataM$TEAM_FIELDING_DP[is.na(ourdataM$TEAM_FIELDING_DP)]<-151

ourdataM$TEAM_BASERUN_CS[is.na(ourdataM$TEAM_BASERUN_CS)]<-50

```

```{r}

cor(ourdataM$TARGET_WINS,ourdataM$TEAM_BASERUN_CS, method="pearson", use = "complete.obs")

cor(ourdataM$TARGET_WINS,ourdataM$TEAM_PITCHING_H, method="pearson", use = "complete.obs")

cor(ourdataM$TARGET_WINS,ourdataM$TEAM_PITCHING_BB, method="pearson", use = "complete.obs")

cor(ourdataM$TARGET_WINS,ourdataM$TEAM_PITCHING_SO, method="pearson", use = "complete.obs")

cor(ourdataM$TARGET_WINS,ourdataM$TEAM_FIELDING_DP, method="pearson", use = "complete.obs")


```

```{r}

#For 4 variables correlation still does not make sense, corrupted variables; we will NA them

ourdataM$TEAM_PITCHING_BB<-NA

ourdataM$TEAM_PITCHING_H<-NA

ourdataM$TEAM_PITCHING_SO<-NA

ourdataM$TEAM_FIELDING_DP<-NA

```

```{r}

head(ourdataM)

```

```{r}

library(car)

ourLM <- lm(TARGET_WINS~ TEAM_BATTING_H+TEAM_BATTING_2B+TEAM_BATTING_3B+TEAM_BATTING_BB+TEAM_BASERUN_CS+TEAM_BASERUN_SB+TEAM_FIELDING_E, data=ourdataM)

summary(ourLM)


```
```{r}

plot(cooks.distance(ourLM))

plot(ourLM)

```




```{r}


ourinf<-influence(ourLM)

sum(ourinf$hat)

```
```{r}

#install.packages("faraway")

library(faraway)

halfnorm(lm.influence(ourLM)$hat,labs=row.names(ourdataM),ylab="Leverages")

```

```{r}

ourLMsum<-summary(ourLM)

ourLMsum$sig

stud<-residuals(ourLM)/(ourLMsum$sig*sqrt(1-ourinf$hat))

qqnorm(stud)

abline(0,1)

```
```{r}

dim(ourdataM)

jack<-rstudent(ourLM)

jack[which.max(abs(jack))]

# 2n/(n-1-p)

qt(.05/(1421*2),1421-1-8)

# no outliers

```



```{r}

ourdataM[rownames(ourdataM)=='183'|rownames(ourdataM)=='1409',]<-NA

ourdataM<-ourdataM[!is.na(ourdataM$TARGET_WINS),]

```



```{r}

LMTBH <- lm(TARGET_WINS~ TEAM_BATTING_H, data=ourdataM)

summary(LMTBH)

with(ourdataM,plot(TEAM_BATTING_H,TARGET_WINS))
 
abline(LMTBH)

LMTBH_R = rstandard(LMTBH)

plot(ourdataM$TEAM_BATTING_H, LMTBH_R, ylab="Standardized Residuals", xlab="BATTING HITS", main="WINS") 
abline(0, 0)         

```



```{r}

LMTB2B <- lm(TARGET_WINS~ TEAM_BATTING_2B, data=ourdataM)

summary(LMTB2B)

with(ourdataM,plot(TEAM_BATTING_2B,TARGET_WINS))
abline(LMTB2B)

LMTB2B_R = rstandard(LMTB2B)

plot(ourdataM$TEAM_BATTING_2B, LMTB2B_R, ylab="Residuals", xlab="BATTING HITS 2 BASE", main="WINS") 
abline(0, 0)  

```



```{r}

# QUDRATIC ?

LMTB3B <- lm(TARGET_WINS~ TEAM_BATTING_3B, data=ourdataM)

summary(LMTB3B)

with(ourdataM,plot(TEAM_BATTING_3B,TARGET_WINS))
abline(LMTB3B)

LMTB3B_R = rstandard(LMTB3B)

plot(ourdataM$TEAM_BATTING_3B, LMTB3B_R, ylab="Residuals", xlab="BATTING HITS 3 BASE", main="WINS") 
abline(0, 0)  

```



```{r}

LMTBHR <- lm(TARGET_WINS~ TEAM_BATTING_HR, data=ourdataM)

summary(LMTBHR)

with(ourdataM,plot(TEAM_BATTING_HR,TARGET_WINS))
abline(LMTBHR)

LMTBHR_R = rstandard(LMTBHR)

plot(ourdataM$TEAM_BATTING_HR, LMTBHR_R, ylab="Residuals", xlab="BATTING HITS HOME RUNS", main="WINS") 
abline(0, 0) 

```



```{r}

LMTBBB <- lm(TARGET_WINS~ TEAM_BATTING_BB, data=ourdataM)

summary(LMTBBB)

with(ourdataM,plot(TEAM_BATTING_BB,TARGET_WINS))
abline(LMTBBB)

LMTBBB_R = rstandard(LMTBBB)

plot(ourdataM$TEAM_BATTING_BB, LMTBHR_R, ylab="Residuals", xlab="BATTING HITS HOME RUNS", main="WINS") 
abline(0, 0) 

```



```{r}

LMTBSB <- lm(TARGET_WINS~ TEAM_BASERUN_SB, data=ourdataM)

summary(LMTBSB)

with(ourdataM,plot(TEAM_BASERUN_SB,TARGET_WINS))
abline(LMTBSB)

LMTBSB_R = rstandard(LMTBSB)

plot(ourdataM$TEAM_BASERUN_SB, LMTBSB_R, ylab="Residuals", xlab="BATTING SB RUNS", main="WINS") 
abline(0, 0) 


```



```{r}

LMTBFE <- lm(TARGET_WINS~ TEAM_FIELDING_E, data=ourdataM)

summary(LMTBFE)

with(ourdataM,plot(TEAM_FIELDING_E,TARGET_WINS))
abline(LMTBFE)

LMTBFE_R = rstandard(LMTBFE)

plot(ourdataM$TEAM_FIELDING_E, LMTBFE_R, ylab="Residuals", xlab="FIELD ERRORS", main="WINS") 
abline(0, 0) 

head(ourdataM)

```

# Build Model


```{r}

powerTransform(cbind(ourdataM$TARGET_WINS,ourdataM$TEAM_BATTING_H,ourdataM$TEAM_BATTING_2B,ourdataM$TEAM_BATTING_3B,ourdataM$TEAM_BATTING_HR,ourdataM$TEAM_BATTING_BB,ourdataM$TEAM_BASERUN_CS,ourdataM$TEAM_BASERUN_SB,ourdataM$TEAM_FIELDING_E))

```

```{r}

ourdataM$TARGET_WINS_BC<-ourdataM$TARGET_WINS^1.23

ourdataM$TEAM_BATTING_H_BC<-ourdataM$TEAM_BATTING_H^0.26

ourdataM$TEAM_BATTING_2B_BC<-ourdataM$TEAM_BATTING_2B^0.44

ourdataM$TEAM_BATTING_3B_BC<-ourdataM$TEAM_BATTING_3B^0.34

ourdataM$TEAM_BATTING_HR_BC<-ourdataM$TEAM_BATTING_HR^0.67

ourdataM$TEAM_BATTING_BB_BC<-ourdataM$TEAM_BATTING_BB^0.68

ourdataM$TEAM_BASERUN_CS_BC<-ourdataM$TEAM_BASERUN_CS^0.02

ourdataM$TEAM_BASERUN_SB_BC<-ourdataM$TEAM_BASERUN_SB^0.17

ourdataM$TEAM_FIELDING_E_BC<-ourdataM$TEAM_FIELDING_E^(-0.3)

```


```{r}

ourLM_BC <- lm(TARGET_WINS_BC~ TEAM_BATTING_H_BC+TEAM_BATTING_2B_BC+TEAM_BATTING_3B_BC+TEAM_BATTING_HR_BC+TEAM_BATTING_BB_BC+TEAM_BASERUN_CS_BC+TEAM_BASERUN_SB_BC+TEAM_FIELDING_E_BC, data=ourdataM)

summary(ourLM_BC)

plot(ourLM_BC)

```











